Scaling rates of true polar wander in convecting planets and moons

### 2017-09-13 tpw_rate.tex
Edits after conversation with Bruce.

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
The vector $\mathbf{g}$ is the gravitational acceleration, 
which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
The gravitational potential obeys 
\begin{equation}
\nabla^2 V = 4 \pi G \rho.
\end{equation}
where $G$ is the gravitational constant. 
For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
\label{eq:eos}
\end{equation}
to define the density.
The remaining parameters are defined in Table~\ref{tab:parameters}.
Note that we retain the centrifugal term in the momentum equation, which is normally either neglected or absorbed into a modified pressure
(in the latter case the boundary conditions on $P$ must be modified).
The centrifugal term
is generally small compared to gravitational forces (at least for Earth-like parameters). Consequently, it does  not have a strong influence on the style of convection.
However, this term is critical for determining the size of the rotational bulge. It is also essential for establishing a connection between the linear and angular momentum equations.

@@

\begin{equation}
T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3.
\end{equation}

@@

Returning to the three-dimensional problem we expect $\Lambda_{ij}$ to scale as
\begin{equation}
\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
\label{eq:lambda_estimate}
\end{equation}
Using this estimate in Equation (33) gives
\begin{equation}
\frac{d W}{dt^\prime} = \frac{1}{m} \sin 2 \theta
\end{equation}
Remarkably, $Ra$ and $\Gamma$ have completely dropped from the expression for the rate of TPW.  The time constant $\tau_R$ goes down at high Rayleigh numbers. At the same time, the  coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{d W}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence 
on the Rayleigh number.

@@

Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be.
\citet{kirschvink1997evidence} hypothesized about an event where $\theta$ was approximately 90$^\circ$,
while many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
In this section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.

@@

Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principal axes. Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  The corresponding rotation of the principal axes is defined by the angle $\xi$. A bound on $\xi$ is given by \citep{davis1970rotation}
\begin{equation}
\vert \sin(2 \xi) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
\label{eq:kahan}
\end{equation} 
Small changes in orientation (i.e. small $\xi$) occur when the difference in eigenvalues is large. Conversely, large changes in orientation are possible when differences between the eigenvalues are small. 
This dependence is illustrated schematically in Figure~\ref{fig:perturb}.  Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}. On the hand hand, \citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33)). If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$. In this case the rate of TPW may be small before and after the interchange.

@@

In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principal axes. In effect, the location of the principal axes are set by the location of a small density anomaly. There is a close correspondence with  hypothesized ``inertial interchange true polar wander'' (IITPW) \citep{kirschvink1997evidence}. As noted above, the interchange of the maximum and intermediate axis can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis. An IITPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).  However, the eigengap does not need to be zero for there to be large polar wander events. In addition, wander does not need to be $90^\circ$ after an interchange of maximum and intermediate axes due to the subsequent evolution of density anomalies during the event.  We clarify these points using numerical simulations of convection in a 2D spherical annulus.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
\caption{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.  Center: Time series of spin axis and mismatch angle $\theta$.  Bottom: Time series of TPW rate. When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger. At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event, with $\sim80^\circ$ degrees of polar wander in $\sim$30 Myr. However, there are several other large TPW events which happen when the eigengap is small.}
\label{fig:misfit}
\end{figure*}

@@

We can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch} to obtain an estimate of maximum TPW rates for Earth.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} performed TPW simulations based on Cenozoic plate motion reconstructions, and argued that the convective planform of Earth has been stable for the last few hundred million years. 
\citet{cambiotti2011new} took a similar approach, but allowed for the position of the spin axis to lag the principal axes (i.e. nonzero $\theta$). In that study they found values for $\theta$ of up to 3$^\circ$-7$^\circ$.
On the other hand, we know that there have been large reorganizations of that planform during Earth history,
so this relative stability of the spin-axis may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.

>>>>

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
The vector $\mathbf{g}$ is the gravitational acceleration, 
which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
The gravitational potential obeys 
\begin{equation}
\nabla^2 V = 4 \pi G \rho.
\end{equation}
where $G$ is the gravitational constant. 
For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is approximately constant (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
\label{eq:eos}
\end{equation}
to define the density.
The remaining parameters are defined in Table~\ref{tab:parameters}.
Note that we retain the centrifugal term in the momentum equation, which is normally either neglected or absorbed into a modified pressure
(in the latter case the boundary conditions on $P$ must be modified).
The centrifugal term
is generally small compared to gravitational forces (at least for Earth-like parameters). Consequently, it does  not have a strong influence on the style of convection.
However, this term is critical for determining the size of the rotational bulge. It is also essential for establishing a connection between the linear and angular momentum equations.

@@

\begin{equation}
T_{\text{degree-two}}(t) \sim \frac{1}{N_{\text{modes}}} \sim \left( \frac{d}{R}\right)^3.
\label{eq:degree_two_of_lengthscale}
\end{equation}

@@

Returning to the three-dimensional problem we expect $\Lambda_{ij}$ to scale as
\begin{equation}
\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
\label{eq:lambda_estimate}
\end{equation}
Using this estimate in Equation (33) gives
\begin{equation}
\frac{d W}{dt^\prime} = \frac{1}{m} \sin 2 \theta
\end{equation}
Remarkably, $Ra$ and $\Gamma$ have completely dropped from the expression for the rate of TPW.  The time constant $\tau_R$ goes down at high Rayleigh numbers. At the same time, the  coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
\citet{tsai2007theoretical} arrive at a similar result by a different path.
Their estimate for the ratio of the convective timescale to the timescale for TPW depends only on $m$ (see their Equation~(4)).
That the two results are similar is surprising, as their estimate of $T_\text{degree-two}$ varies linearly with $d/R$,
rather than with $(d/R)^3$ (Equation~\eqref{eq:degree_two_of_lengthscale}).

The cancellation of explicit $\mathrm{Ra}$ dependence is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{d W}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence 
on the Rayleigh number.

@@

Much of the debate about the magnitude of TPW on Earth comes down to the question of how big $\theta$ can be.
\citet{kirschvink1997evidence} hypothesized about an inertial interchange event, where $\theta$ alters by approximately 90$^\circ$,
and can lead to large ($\sim90^\circ$) TPW over time (a process which they termed ``inertial interchange TPW, or IITPW).
Conversely, many geodynamic simulations of recent TPW assume that $\theta$ is small \citep{steinberger1997changes}.
In this section we seek to characterize the maximum value for $\theta$ in terms of the dimensionless parameters.

@@

Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principal axes. Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  The corresponding rotation of the principal axes is defined by the angle $\xi$. A bound on $\xi$ is given by \citep{davis1970rotation}
\begin{equation}
\vert \sin(2 \xi) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
\label{eq:kahan}
\end{equation} 
Small changes in orientation (i.e. small $\xi$) occur when the difference in eigenvalues is large. Conversely, large changes in orientation are possible when differences between the eigenvalues are small. 
This dependence is illustrated schematically in Figure~\ref{fig:perturb}.  Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}. On the other hand, \citet{tsai2007theoretical} point out that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33)). If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$. In this case the rate of TPW may be small before and after the interchange.

@@

In the limit that the eigengap becomes zero, a small perturbation $\delta$ can produce an arbitrary rotation of the principal axes. In effect, the location of the principal axes are set by the location of a small density anomaly. This has a close correspondence with hypothesized IITPW \citep{kirschvink1997evidence}. As noted above, the interchange of the maximum and intermediate axex can produce a large change in $\theta$ because the misalignment angle is suddenly measured from a new axis. The resulting TPW event can produce a large change in the orientation of the rotation axis (potentially $90^{\circ}$ if there are not further perturbations to the $\mathbf{E}$ tensor).  However, the eigengap does not need to be zero for there to be large polar wander events. In addition, wander does not need to be $90^\circ$ after an interchange of maximum and intermediate axes due to the subsequent evolution of density anomalies during the event.  We clarify these points using numerical simulations of convection in a 2D spherical annulus.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
  \caption{Top: Time series of principal moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.  Center: Time series of spin axis and mismatch angle $\theta$.  Bottom: Time series of TPW rate. When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger. At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event, with $\sim80^\circ$ degrees of polar wander in $\sim$30 Myr. However, there are several other large TPW events which happen when the eigengap is small. These results are fairly characteristic of high $\mathrm{Ra}$ convection.}
\label{fig:misfit}
\end{figure*}

@@

We can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch} to obtain an estimate of maximum TPW rates for Earth.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} performed TPW simulations based on Cenozoic plate motion reconstructions, and argued that the convective planform of Earth has been stable for the last few hundred million years. 
\citet{cambiotti2011new} took a similar approach, but allowed for the position of the spin axis to lag the principal axes (i.e. nonzero $\theta$). In that study they found values for $\theta$ of up to 3$^\circ$-7$^\circ$.
On the other hand, we know that there have been large reorganizations of that planform during Earth history,
so this relative stability of the spin-axis may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{\tau_R} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr at a point 90$^\circ$ from the TPW axis (and is comparable to the maximum rates seen in Figure~\ref{fig:misfit}.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.
